Supermartingales and Termination Analysis of Statistical PCF

### 2021-01-02 LICS21/SPCF.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable. 
We say that a term $M$ is \emph{almost-surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid M \text{ terminates at } s\}) = 1$.

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
For example, the term 
\[
(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\[
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
\] 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

@@

%\paragraph{}
This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

>>>>

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable. 
We say that a term $M$ is \emph{almost surely terminating} (AST) just if $\mu_{\entrosp}(\{s \mid M \text{ terminates at } s\}) = 1$.

\iffalse
\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost surely terminating} (PAST) if $\expect{T_M} < \infty$.}
\fi
For example, the term 
\[
(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $\entrosp \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\[
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
\] 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

@@

%\paragraph{}
This definition of almost sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

### 2021-01-02 LICS21/antitone.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

The following theorem does not prove completeness, but is suggestive in that direction:
\begin{restatable}{theorem}{antitoneSeemsComplete}
For any stopping time $T$ which is almost-surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{restatable}

@@

\begin{example}[Higher-order recursion]
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \; n \; (\varphi \; f \; s \; (n - 1))
\oplus_p
f \; (n+1) \; (\varphi \; f \; s \; n)
}
\end{array}
\]
\lo{QUESTION: Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
Then $\Xi \, g$ is AST for $p \geq 1/2$.}
\akr{In order to actually construct a ranking function, it will be necessary to assume a bound on the number of $\tY$-reduction steps in the reduction of $g \underline n \underline m$ too.}
\end{example}

>>>>

The following theorem does not prove completeness, but is suggestive in that direction:
\begin{restatable}{theorem}{antitoneSeemsComplete}
For any stopping time $T$ which is almost surely finite, if $(\calF_n)_n$ is the coarsest filtration to which $T$ is adapted, then there is a supermartingale $(Y_n)_n$ adapted to $(\calF_n)_n$ and an antitone function $\epsilon$ such that $(Y_n)_n$ is an antitone ranking supermartingale with respect to $T$ and $\epsilon$.
\end{restatable}

@@

\begin{example}[Higher-order recursion]
Consider the higher-order function
$\Xi : (\textsf{R} \to \textsf{R} \to \textsf{R}) \to \textsf{R} \to \textsf{R} \to \textsf{R}$
recursively defined\footnote{Inspired by examples in \citep{DBLP:journals/pacmpl/BurnOR18,DBLP:conf/lics/OngW19}.} by
\[
\begin{array}{l}
\Xi := \tY \lambda \varphi \, f^{\textsf{R} \to \textsf{R} \to \textsf{R}} \, s^\textsf{R} \, n^\textsf{R} . \\
\quad\tif{n \leq 0}{s}{
f \; n \; (\varphi \; f \; s \; (n - 1))
\oplus_p
f \; (n+1) \; (\varphi \; f \; s \; n)
}
\end{array}
\]
\lo{QUESTION: Let $g : \textsf{R} \to \textsf{R} \to \textsf{R}$ be a function that terminates on all arguments.
Then $\Xi \, g$ is AST for $p \geq 1/2$.}
\akr{In order to actually construct a ranking function, it will be necessary to assume a bound on the number of $\tY$-reduction steps in the reduction of $g \, \underline n \, \underline m$ too.}

\akr{I suspect that the term you actually had in mind for the higher-order recursion example was different from what you wrote, because you said that it terminates for $p \geq 1/2$, but it actually terminates for all $p > 0$. Passing $n+1$ as an argument to $f$ doesn't actually make the recursion any longer. Did you mean 
\[
\begin{array}{l}
\tY \lambda \varphi f s n. \\
\qquad \tif{n \leq 0}{s}{ f \, n \, (\varphi \, f \, s \, (n - 1)) \oplus_p f \, n \, (\varphi \, f \, s \, (n+1))}
\end{array}
\]}
\end{example}

### 2021-01-02 LICS21/confluent.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to. Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far, or considering an alternative reduction order may be simpler or more intuitive.

@@

Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$. If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}. 
The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ inaffected.
\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?}

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
%\lo{On first reading, I struggled to parse the following clauses.}
\begin{definition}
\label{def:more general red}
The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
\begin{align*}
  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$}\\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
\end{align*}
In each of these cases, $M \mid \alpha$ is the \emph{redex}, and reduction takes place at $\alpha$.
\end{definition}

@@

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of traces where $A[\tsample]$ terminates would be
\[\begin{array}{l}
\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid \\
\qquad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
\qquad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
\end{array}\] 
This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. This set is unmeasurable, therefore the termination probability would not even be well-defined. Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet. For this particular term, $Sk(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
\begin{align*}
&\{s \in I^{L_s(A[\tsample])} \mid \\
&\qquad s(Sk(A[\tsample]), Sk(A[\underline 0]), \ldots, Sk(B), \mathsf{if}_1) > 0.5 \},
\end{align*}
which is measurable. Skeletal reduction sequences still contain just as much informationas reduction sequences of terms, however, on which reductions took place.
\end{example}

>>>>

When proving almost sure termination in this way, it is necessary to consider which terms a given term may reduce to. Sometimes however, the reduction that the programmer has in mind may not be strictly the call-by-value order defined so far; or considering an alternative reduction order may be simpler or more intuitive.

@@

Two subterms $N_1$ and $N_2$ of a term $M$, corresponding to positions $\alpha_1$ and $\alpha_2$, can overlap in a few different ways. 
If $\alpha_1$ is an initial segment (i.e.~prefix) of $\alpha_2$ (written as $\alpha_1 \leq \alpha_2$), then $N_2$ is also a subterm of $N_1$. If neither $\alpha_1 \leq \alpha_2$ nor $\alpha_1 \geq \alpha_2$, the positions are said to be \emph{disjoint}. 
The notion of disjointness is mostly relevant in that if $\alpha_1$ and $\alpha_2$ are disjoint, performing a substitution at $\alpha_1$ will leave the subterm at $\alpha_2$ unaffected.
\lo{The preceding sentence is unclear. Typo?} \akr{It wasn't a typo, but I've made it a little more explicit now. Is this clearer?} \lo{Yes, it's clear now.}

With this notation, a more general reduction relation $\to$ can be defined. \akr{It would almost work to have one of the neater versions of $\tY$-reduction here.}
%\lo{On first reading, I struggled to parse the following clauses.}
\begin{definition}
\label{def:more general red}
The binary relation $\to$ is defined by the following rules, each is conditional on a redex occurring at position $\alpha$ in the term $M$:
\begin{align*}
  \text{if } M \mid \alpha = (\lambda x.N) V,\ & M \to M[N[V/x]/\alpha] \\
  \text{if } M \mid \alpha = \underline f (\underline r_1, \dots , \underline r_n),\ & M \to M[\underline{f(r_1,\dots,r_n)}/\alpha] \\
  \text{if } M \mid \alpha = \tY \lambda x. N,\ & M \to M[\lambda z. N[(\tY \lambda x. N)/x] z/\alpha]\\ \text{where }z&\text{ is not free in $N$}\\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_1/\alpha] \text{ where }r < 0 \\
  \text{if } M \mid \alpha = \tif{\underline r < 0}{N_1}{N_2},\ & M \to M[N_2/\alpha] \text{ where }r \geq 0 \\
  \text{if } M \mid \alpha = \tsample \text{ and $\lambda$ does}&\text{ not occur after $@_2$ or $\tY$ in $\alpha$},\\ & M \to M[\underline r/\alpha] \text{ where } r \in [0,1] \\
  \text{if } M \mid \alpha = \tscore(\underline r),\ & M \to M[\underline r/\alpha].
\end{align*}
In each of these cases, $M \mid \alpha$ is the \emph{redex}, and the reduction takes place at $\alpha$.
\end{definition}

@@

\begin{example}[Labelling samples by terms]
Consider the terms
\begin{align*}
A[x] &= \tif{\tif{x>0}{I}{I} \, \tsample - 0.5 > 0}{0}{\Omega}\\
B &= \tif{\tsample - 0.5 > 0}{0}{\Omega}
\end{align*}
If terms rather than skeletons were used to label samples, the set of traces where $A[\tsample]$ terminates would be
\[\begin{array}{l}
\bigcup_{r \in [0,1]} \{s \in I^{L_s(A[\tsample])} \mid \\
\qquad\qquad\qquad\quad s([A[\tsample]], \mathsf{if}_1;\underline{-}_1;@_1;\mathsf{if}_1) = r, \\
\qquad\qquad\qquad\quad s([A[\tsample], A[\underline r], \ldots, B, \mathsf{if}_1) > 0.5 \}.
\end{array}\] 
This is a rather unwieldy expression, but the crucial part is that $r$ occurs twice in the conditions on $s$: once as the value a sample must take, and once in the location of a sample. This set is unmeasurable, therefore the termination probability would not even be well-defined. Labelling samples by skeletons instead, this problem does not occur because there are only countably many skeleton, and at each step in a reduction sequence, only finitely many could have occurred yet. For this particular term, $Sk(A[\underline r])$ does not depend on the value of $r$, therefore the set where it terminates becomes simply
\begin{align*}
&\{s \in I^{L_s(A[\tsample])} \mid \\
&\qquad s(Sk(A[\tsample]), Sk(A[\underline 0]), \ldots, Sk(B), \mathsf{if}_1) > 0.5 \},
\end{align*}
which is measurable. Skeletal reduction sequences still contain just as much informationas reduction sequences of terms, however, on which reductions took place.
\qed
\end{example}

### 2021-01-02 LICS21/ranking.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

Although rankability implies almost-sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
\label{ex:0 probability reachable}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.

@@

\subsection{Sparse ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost-sure termination, so it isn't useful for proving it.

@@

\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
\fi
Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$. The meanings of the substitutions and conditions that label some of the reductions in the diagram should be self-explanatory.

>>>>

Although rankability implies almost sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
\label{ex:0 probability reachable}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.

@@

\subsection{Sparse ranking functions}
Even in the case of reasonable simple terms, explicitly constructing a ranking function would be a lot of work, and \Cref{thm:minimal} makes even stronger assumptions than almost sure termination, so it isn't useful for proving it.

@@

\lo{The graph can be simplified somewhat if we redefine general recursion using a fixpoint operator $\tY_{f, n}\Theta$ with redex rule as per \Cref{eq:letrec redex}: see \Cref{fig:geometric distribution 2}.
But (especially in view of the vast investment in the confluent trace semantics) it is perhaps too late to change things now?}
\fi
Despite its simplicity, its $Rch$ contains all the terms in \Cref{fig:geometric distribution}, for each $i \in \mathbb N, r \in I$. The meanings of the substitutions \changed[lo]{(to be applied to the contractum)} and conditions that label some of the reductions in the diagram should be self-explanatory.

### 2021-01-02 LICS21/supermartingales-proofs.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

It is straightforward to see that for every trace $s \in S$, there is a reduction sequence of $\transform{M}$ that simulates (an initial subsequene of) the reduction of $M$ under $s$;
and the simulation can be made lockstep.

>>>>

It is straightforward to see that for every trace $s \in \entrosp$, there is a reduction sequence of $\transform{M}$ that simulates (an initial subsequene of) the reduction of $M$ under $s$;
and the simulation can be made lockstep.

### 2021-01-02 LICS21/supermartingales.tex
Merge branch 'master' of https://github.com/4Denthusiast/prob-programming-termpaper
# Conflicts:
#	LICS21/ranking.tex

almost-sure -> almost sure (for consistency)

Plainly
\(
f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
\)
$\mu$-almost-surely.
%(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
Therefore 
\[
\expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
= 
\expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\]
\begin{align}
& \expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
\nonumber \\
& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\label{eqn:linear cond ex} 
& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
\label{eqn:inde cond exp}
& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
\end{align}
\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
%\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).

@@

\begin{theorem}[Soundness of rankability] \label{thm:rankable implies termination}
\begin{enumerate}
\item If a closed SPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost-sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.

\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost-sure terminating (PAST).
\end{enumerate}
\end{theorem}

>>>>

Plainly
\(
f(M_{n+1}) = %\sum_{i=1}^3 f_{i, n+1}(M).
f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M),
\)
$\mu$-almost surely.
%(Unless otherwise stated, all equations and inequations between random variables are assumed to hold only $\mu$-a.s.)
Therefore 
\[
\expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
= 
\expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\]
\begin{align}
& \expect{f(M_{n+1}) \mid \mathcal{F}_n} \nonumber \\
& = \expect{f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)  \mid \mathcal{F}_n} 
\nonumber \\
& = \expect{f_{1, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{2, n+1}(M) \mid \mathcal{F}_n} + \expect{f_{3, n+1}(M) \mid \mathcal{F}_n} 
\label{eqn:linear cond ex} 
& = f_{1, n+1}(M) + f_{2, n+1}(M) + f_{3, n+1}(M)
\label{eqn:inde cond exp}
& = f_{1, n+1}(M) + \int_I f() + f_{3, n+1}(M) \label{eqn:inde cond exp} \\
\end{align}
\Cref{eqn:linear cond ex} follows from the linearity of conditional expectation. 
%\Cref{eqn:inde cond exp} is justified because $f_{i, n+1}(M)$ is $\mathcal{F}_n$-measurable (\Cref{lem:inde}).

@@

\begin{theorem}[Soundness of rankability] \label{thm:rankable implies termination}
\begin{enumerate}
\item If a closed SPCF term $M$ is rankable, then $M$ is AST (equivalently, $T^{\tY}_M < \infty$ almost surely), and $M$ is $\tY$-positively almost sure terminating ($\tY$-PAST) i.e.~$\expect{T^{\tY}_M} < \infty$.

\item If a closed SPCF term $M$ is strictly rankable, then $\expect{T_M} < \infty$ i.e.~$M$ is positively almost sure terminating (PAST).
\end{enumerate}
\end{theorem}
